Resonant Scattering in a Uniform Sphere with Large Optical Depth

### 2021-10-12 ms.tex
Adding last of Shane's edits to time dependent section

\subsection{Comparison to Monte Carlo}
The Monte Carlo method is used to solve the transfer equation numerically. This method is valid at arbitrarily large $\tau_0$, being restricted only by its computational demand which grows proportionally to the number of photons used and $\tau_0$. For each simulation, a total of ${\sim}10^6$ photon packets are initialized at a monochromatic source frequency $x_s$ and are allowed to propagate through the sphere until escaping, at which point their positions, outgoing angles, and escape frequencies are stored. A constant temperature of $T=10,000\ \rm K$ is set for the gas. Frequency redistribution is calculated at each scattering, including the effects of recoil, to obtain the spectrum at the surface of the spherical simulation domain. In the comparisons shown in this section, the raw photon data is binned in frequency to obtain spectra. 64 bins are used to ensure good statistics and minimal noise in each bin. Further details of the Monte Carlo implementation are discussed in \cite{2017ApJ...851..150H}.

@@

We now compare each of the previously-discussed solutions for surface flux with Monte Carlo simulations. We return to the Doppler units $x$ for the photon frequency rather than the $\sigma$ units used in the solutions of the transfer equation. The spectrum $P(x)$ is defined as the specific luminosity at the surface divided by the source luminosity, or
\be
P(x) = \frac{16\pi^2R^2H(R, x)\Delta}{L}.
\ee
This is normalized such that $\int P(x)dx = 1$. Since $H(r, x)$ is per $d\nu$, a factor of $\Delta$ gives the expression the correct units.

In Figure \ref{fig:sol_mc_residual_0}, the Monte Carlo escape probability as a function of frequency is shown with that of the solutions $H_{\rm d}$, $H_0$, and $H_{\rm 0 + bc} = H_0 + H_{\rm bc}$ for an optical depth of $\tau_0 = 10^7$ and photons emitted at line center $\rm x_s = 0$.  The $H_{\rm bc}$ term is negative at the peak of the spectrum and positive in the line wing such that, when added to the solution $H_0$ derived by \citet{2006ApJ...649...14D}, it adequately corrects for the apparent excess of flux in the peaks of the spectrum. The solution with the correct frequency-dependent boundary condition enforced, $H_{0 + \rm bc}$, has considerably lower residuals to Monte Carlo than the other solutions, especially in the line wing. The boundary term corrects the slight deficit of $H_0$ in the line wings, further improving agreement with the numerical result. Note that the residuals to the $H_0$ solution are a close match to the $H_{\rm bc}$ term, since the Monte Carlo represents the ``true'' solution, $H$, and $H_{\rm bc} = H - H_0$. It is evident that the divergent solution $H_{\rm d}$ fails in the line wings.

@@

The size of $H_{\rm bc}$ is dependent on $\tau_0$, which also dictates the x-position of the peak of spectrum according to Eq. \ref{eq:tau_peak_scaling}. This term is significant even at $\tau_0 {\sim} 10^7$ where $H_0$ solution is expected to perform its best, i.e., photons are pushed further out into the wing and thus the simplifying assumptions made in the solution of the differential equation are a better approximation.

In Figure \ref{fig:sol_mc_tau} we show the solutions alongside Monte Carlo for $\tau_0=10^5, 10^6$, and $10^7$. From Eq. \ref{eq:hbc_scaling}, the size of the term $H_{\rm bc}$ should become smaller with larger optical depths, following a $(a\tau_0)^{-1/3}$ scaling. Indeed, agreement between Eq. \ref{eq:H0surf} and the Monte Carlo points in Fig. \ref{fig:sol_mc_tau} improves as $\tau_0$ increases, with $H_{\rm bc}$ providing a fractionally smaller correction to $H_0$. One factor of $(a\tau_0)^{1/3}$ has been scaled out of the x-axis such that the peaks of the distributions are vertically aligned. This scaling has also been applied to the y-axis to preserve normalization of the escape probability. At lower $\tau_0$, the scattering of photons within the Doppler core of the line becomes important, yet our solution does not include this effect. The effects of line core scattering can be seen in the Monte Carlo data, however, as the peaks of the distribution shift substantially outward in frequency from the line core for $\tau_0 = 10^5$.

@@

Figure \ref{fig:sol_mc_xinit} shows calculations performed for three initial photon frequencies $\rm x_s = 0.0, 6.0$, and $12.0$ for a line center optical depth of $\tau_0=10^7$. The asymmetry of the spectrum is slight for $\rm x_s = 6.0$ where the cross section is still relatively large, but the increase in mean free path is much more substantial when photons are initialized with $\rm x_s=12.0$, further away from the line core. It is seen here that the difference between the Monte Carlo data and $H_0$ becomes larger as $\rm x_s$ increases. Thus, for sufficiently large $\tau_0$ and emission away from line center ($x_s \neq 0$), the importance of enforcing frequency-dependent boundary conditions at the surface of the sphere grows.

Figure \ref{fig:sol_mc_xinit_lowtau} shows emission away from line center at the same values of $x_s$ as in Figure \ref{fig:sol_mc_xinit}, but for $\tau_0=10^6$ rather than $10^7$. The difference between the left and right side of the escaping spectrum is substantial at lower optical depths, since a change in $x_s$ of 6 is a larger fraction of the spectral peak location $(a\tau_0)^{1/3}$ for lower $\tau_0$. It is clear from the figure that as $x_s$ becomes further from the line core, the spectrum approaches a delta function in frequency. Additionally, since the material is optically thin here, we expect there to be a stronger disagreement between the Monte Carlo and the analytic solution as derivations assumed large optical depths.

@@

Figure \ref{fig:xinit} shows quantitative examination of how the boundary term $H_{\rm bc}$ scales with both $\tau_0$ and $x_s$, motivated by the previous figures' demonstration that $H_0$ appears to become more accurate as $\tau_0$ increases. In this figure, the photon initialization frequency $x_s$ is shifted by integer multiples of $\tau_0$ in $\sigma$ frequency units (Eq. \ref{eq:change_of_variables}) such that the source, now written $\sigma_s$, falls near the peak of the spectrum for each $\tau_0$. For clarity, only the $x > 0$ side of the spectrum is shown. From Eq. \ref{eq:hbc_scaling}, it is expected that the size of $H_{\rm bc}$ should become smaller with larger optical depths following a $(a\tau_0)^{-1/3}$ scaling as compared to $H_0$. This factor has been scaled out of the figure such that solutions for different optical depths and the same source frequency $\sigma_s$ should show close agreement in scale on the figure's vertical axis if the relation holds. Indeed, the scaled solutions converge as the optical depth becomes larger, indicating general agreement with the $(a\tau_0)^{-1/3}$ scaling. The remaining discrepancy present in the vertical axis for fixed $\sigma_s$ is due to the width of the Doppler core with respect to the peak of the spectrum; at lower $\tau_0$, this causes the line profile approximation in the wing (Eq. \ref{eq:line_profile_wing}) to break down. From this, we conclude that the errors introduced by the incorrect separation of variables in \cite{1973MNRAS.162...43H}, \cite{1990ApJ...350..216N}, \cite{2006ApJ...649...14D} and others are proportional to $(a\tau_0)^{-1/3}$.

@@

\acknowledgments
\restartappendixnumbering

>>>>

\subsection{Comparison to Monte Carlo}
The Monte Carlo method is used to solve the transfer equation numerically. This method is valid at arbitrarily large $\tau_0$, being restricted only by its computational demand which grows proportionally to the number of photons used and $\tau_0$. For each simulation, a total of ${\sim}10^6$ photon packets are initialized at a monochromatic source frequency $x_s$ and are allowed to propagate through the sphere until escaping, at which point their positions, outgoing angles, and escape frequencies are tabulated to obtain the spectrum at the surface of the spherical simulation domain. A constant temperature of $T=10^4\ \rm K$ is set for the gas. Frequency redistribution is calculated at each scattering, including the effects of recoil. In the comparisons shown in this section, the raw photon data is binned in frequency to obtain spectra. Further details of the Monte Carlo implementation are discussed in \cite{2017ApJ...851..150H}.

@@

We now compare each of the previously-discussed solutions for surface flux to the Monte Carlo results. We return to the Doppler units $x$ for the photon frequency rather than the $\sigma$ units used in the solutions of the transfer equation. The spectrum $P(x)$ is defined as the specific luminosity at the surface divided by the source luminosity, or
\be
P(x) = \frac{16\pi^2R^2H(R, x)\Delta}{L}.
\ee
This is normalized such that $\int P(x)dx = 1$. Since $H(r, x)$ is per $d\nu$, a factor of $\Delta$ gives the expression the correct units.

In Figure \ref{fig:sol_mc_residual_0}, the Monte Carlo escape probability as a function of frequency is shown with that of the solutions $H_{\rm d}$, $H_0$, and $H_{\rm 0 + bc} = H_0 + H_{\rm bc}$ for an optical depth of $\tau_0 = 10^7$ and photons emitted at line center $\rm x_s = 0$.  The $H_{\rm bc}$ term is negative at the peak of the spectrum and positive in the line wing such that, when added to $H_0$, it adequately corrects for the apparent excess of flux in the peaks of the spectrum. The solution with the correct frequency-dependent boundary condition enforced, $H_{0 + \rm bc}$, has considerably lower residuals to Monte Carlo results than the other solutions, especially in the line wing. The boundary term corrects the slight deficit of $H_0$ in the line wings, further improving agreement with the numerical result. Note that the residuals to the $H_0$ solution are a close match to the $H_{\rm bc}$ term, since the Monte Carlo represents the ``true'' solution, $H$, and $H_{\rm bc} = H - H_0$. It is evident that the divergent solution $H_{\rm d}$ fails in the line wings.

@@

The size of $H_{\rm bc}$ is dependent on $\tau_0$, which also dictates the x-position of the peak of spectrum according to Eq. \ref{eq:tau_peak_scaling}. This term is significant even at $\tau_0 {\sim} 10^7$ where the $H_0$ solution is expected to perform best, i.e., photons are pushed further out into the wing where the simplifying assumptions made in the solution of the differential equation are a better approximation.

In Figure \ref{fig:sol_mc_tau} we show the solutions alongside Monte Carlo for $\tau_0=10^5, 10^6$, and $10^7$. From Eq. \ref{eq:hbc_scaling}, the size of the term $H_{\rm bc}$ should become smaller with larger optical depths, following a $(a\tau_0)^{-1/3}$ scaling. Indeed, agreement between Eq. \ref{eq:H0surf} and the Monte Carlo points in Fig. \ref{fig:sol_mc_tau} improves as $\tau_0$ increases, with $H_{\rm bc}$ providing a fractionally smaller correction to $H_0$. One factor of $(a\tau_0)^{1/3}$ has been scaled out of the x-axis such that the peaks of the distributions are vertically aligned. This scaling has also been applied to the y-axis to preserve normalization of the escape probability. At lower $\tau_0$, the scattering of photons within the Doppler core of the line becomes important, but our solution does not include this effect. The effects of line core scattering can be seen in the Monte Carlo data, however, as the peaks of the distribution shift substantially outward in frequency from the line core for $\tau_0 = 10^5$.

@@

Figure \ref{fig:sol_mc_xinit} shows calculations performed for three initial photon frequencies $\rm x_s = 0.0, 6.0$, and $12.0$ for a line center optical depth of $\tau_0=10^7$. The asymmetry of the spectrum is slight for $\rm x_s = 6.0$ where the cross section is still relatively large, but the increase in mean free path is much more substantial when photons are initialized with $\rm x_s=12.0$, outside the line core. It is seen here that the difference between the Monte Carlo data and $H_0$ becomes larger as $\rm x_s$ increases. Thus, for sufficiently large $\tau_0$ and emission away from line center ($x_s \neq 0$), the importance of enforcing frequency-dependent boundary conditions at the surface of the sphere grows.

Figure \ref{fig:sol_mc_xinit_lowtau} shows emission away from line center at the same values of $x_s$ as in Figure \ref{fig:sol_mc_xinit}, but for $\tau_0=10^6$ rather than $10^7$. The difference between the left and right side of the escaping spectrum is substantial at lower optical depths, since a change in $x_s$ of 6 is a larger fraction of the spectral peak location $(a\tau_0)^{1/3}$ for lower $\tau_0$. It is clear from the figure that as $x_s$ moves further from the line core, the spectrum becomes more strongly peaked in frequency. Additionally, since the material is increasingly optically thin in the wing, we expect there to be a stronger disagreement between the Monte Carlo and the analytic solution as our derivation assumed large optical depths.

@@

Figure \ref{fig:xinit} shows how the boundary term $H_{\rm bc}$ scales with both $\tau_0$ and $x_s$.%, motivated by the previous figures' demonstration that $H_0$ appears to become more accurate as $\tau_0$ increases. 
In this figure, $x_s$ is shifted by integer multiples of $\tau_0$ in $\sigma$ frequency units (Eq. \ref{eq:change_of_variables}) such that the source, now written $\sigma_s$, falls near the peak of the spectrum for each $\tau_0$. For clarity, only the $x > 0$ side of the spectrum is shown. From Eq. \ref{eq:hbc_scaling}, it is expected that the fractional size of $H_{\rm bc}$ relative to $H_0$ should become smaller with larger optical depths following $(a\tau_0)^{-1/3}$. This factor has been scaled out of the figure such that solutions for different optical depths and the same source frequency $\sigma_s$ should show close agreement in scale on the figure's vertical axis if the relation holds. Indeed, the scaled solutions converge as the optical depth becomes larger, indicating general agreement with the $(a\tau_0)^{-1/3}$ scaling. The remaining discrepancy present in the vertical axis for fixed $\sigma_s$ is due to the width of the Doppler core with respect to the peak of the spectrum; at lower $\tau_0$, this causes our line profile approximation (Eq. \ref{eq:line_profile_wing}) to break down. From this, we conclude that the errors introduced by the incorrect separation of variables in \cite{1973MNRAS.162...43H}, \cite{1990ApJ...350..216N}, \cite{2006ApJ...649...14D} and others are proportional to $(a\tau_0)^{-1/3}$.

@@

\acknowledgments

NASA ATP grant 80NSSC18K0696, ``Exoplanetary MHD Outflows Driven by EUV Heating, Lyman alpha Radiation Forces and Stellar Tides".
\restartappendixnumbering
